Topics in Statistical Theory --- Example Sheet 1

### 2020-10-20 modern_statistical_methods/msm_ES1.tex
updated notes and worked on msm exercises

@@

	and therefore $K^{(1)} \circ K^{(2)} = \sum_{m, \ell} \lambda_m \mu_\ell \qty[(P_m \circ Q_\ell)(P_m \circ Q_\ell)\T]$, which is a nonnegative combination of positive semi-definite matrices and therefore positive semi-definite. 
\end{enumerate}

\end{proof}
\end{document}

>>>>

\newcommand\XX{\Cal X}

@@

	and therefore $K^{(1)} \circ K^{(2)} = \sum_{m, \ell} \lambda_m \mu_\ell \qty[(P_m \circ Q_\ell)(P_m \circ Q_\ell)\T]$, which is a nonnegative combination of positive semi-definite matrices and therefore positive semi-definite. 
\end{enumerate}
\end{proof}

\begin{question}
	Let $\XX \ceq \qty{x \in \RR^d : \norm{x} < 1}$. Show that $k(x, x') \ceq (1 - x\T x')^{-\alpha}$ defined on $\XX \times \XX$, where $\alpha > 0$, is a kernel.
\end{question}

\begin{proof}
	Note by Cauchy-Schwarz we have for $x, x' \in \XX$ that $\abs{x\T x'} \leq \norm{x} \norm{x'} < 1$, so $k$ is well-defined.

	Symmetry of $k$ is clear. Define the function 
	\[
	\phi \colon (-\infty, 1) \to \RR \colon x \mapsto (1-x)^{-\alpha}. 
	\]
	Then $\phi$ is continuous, and we can compute its Taylor series around 0. It is computed that 
	\[
	\phi^{(k)}(x) = (1-x)^{-\alpha - k} \prod_{j=0}^{k-1}(\alpha + j) \implies \phi^{(k)}(0) = \prod_{j=0}^{k-1} (\alpha + j). 	\]
	Therefore, the Taylor series of $\phi$ is given by
	\[
	\sum_{k=0}^\infty \qty(\prod_{j=0}^{k-1} (\alpha + j)) \frac{x^k}{k!}, 
	\]
	and its radius of convergence is given by
	\[
	\lim_{k\to\infty} \frac{\frac{1}{k!}\prod_{j=0}^{k-1} (\alpha + j)}{\frac{1}{(k+1)!} \prod_{j=0}^k (\alpha + j)} = \lim_{k\to\infty} \frac{k}{\alpha + k} = 1,
	\]
	so $\phi$ agrees with its Taylor series on $(-1, 1)$. By Cauchy-Schwarz, for $x, x' \in \XX$ we have $x_i\T x_j \leq \norm{x_i} \norm{x_j} < 1$, so we can write for $x, x' \in \XX$
	\[
	k(x, x') = \phi(x\T x') = \sum_{k=0}^\infty \qty(\prod_{j=0}^{k-1} (\alpha + j)) \frac{(x\T x')^k}{k!}.
	\]
	By proposition 4, every term in the sum is a kernel, and therefore the limit is a kernel as well. 
\end{proof}

\begin{question}
	Suppose we have a matrix of predictors $X \in \RR^{n \times p}$ where $p \gg n$. Explain how to obtain the fitted values of the following ridge regression using the kernel trick:
	\begin{gather*}
		\text{Minimise over $\beta \in \RR^p$, $\theta \in \RR^{p(p-1)/2}$, $\gamma \in \RR^p$,} \\
		\sum_{i=1}^n \qty(Y_i - \sum_{k=1}^p X_{ik}\beta_k - \sum_{k=1}^p \sum_{j=1}^{k-1} X_{ik} X_{ij} \theta_{jk} - \sum_{k=1}^p X_{ik}^2\gamma_k)^2 + \lambda_1 \norm{\beta}^2 + \lambda_2 \norm{\theta}^2 + \lambda_3\norm{\gamma}^2. 
	\end{gather*}
Note that we have indexed $\theta$ with two numbers for convenience. 
\end{question}

\begin{proof}
	Our linear model has predictors
	\[
	\qty{X_{ik} \mid 1 \leq k \leq p} \cup \qty{X_{ij} X_{ik} \mid 1 \leq j < k \leq p} \cup \qty{X_{ik}^2 \mid 1 \leq k \leq p}
	\]
	for $i = 1, \dotsc, k$. We want to assume that $\lambda_1 = \lambda_2 = \lambda_3$: to this end, define $\xi = \frac{\lambda_1}{\lambda_2}$ and $\eta = \frac{\lambda_3}{\lambda_1}$, and replace $\theta$ by $\xi\theta$ and $\gamma$ by $\eta\gamma$. This means we will have to replace our predictors by
	\[
	Z_i \ceq 
	(X_{ik} \mid 1 \leq k \leq p)\T \cup (\xi X_{ij} X_{ik} \mid 1 \leq j < k \leq p)\T \cup (\gamma X_{ik}^2 \mid 1 \leq k \leq p)\T
	\]
	We use the kernel trick: note that
	\begin{align*}
	\ang{Z_i, Z_j} &= \sum_k X_{ik}X_{jk} + \xi^2 \sum_{k < \ell} X_{ik} X_{i\ell} X_{jk} X_{j\ell} + \gamma^2 \sum_k X_{ik}^2 X_{jk}^2 \\
	&= \qty(\frac{1}{\sqrt 2\xi} + \frac{\xi}{\sqrt 2}\sum_k X_{ik}X_{jk})^2 + (\gamma^2 - \frac{\xi^2}{2})\sum_k X_{ik}^2 X_{jk}^2 - \frac{1}{2\xi^2}.
	\end{align*}
	We have therefore rewritten $\ang{Z_i, Z_j}$ into something that can be computed in $O(p)$, which is exactly the aim of the kernel trick.
\end{proof}

\begin{question}
	Let $\hat\alpha$ be a minimiser of $\norm{Y - K\alpha}^2 + \lambda \alpha\T K\alpha$, with $K$ being a kernel matrix as usual. Show that $K\hat\alpha = K(K + \lambda I)^{-1} Y$. 
\end{question}

\begin{solution}
	Suppose $\hat\alpha$ minimises $Q(\alpha) \ceq \norm{Y - K\alpha}^2 + \lambda \alpha\T K\alpha$. Then we have
	\begin{align*}
		0 = \nabla_\alpha Q(\hat\alpha) &= -2 K (Y - K\hat\alpha) + 2\lambda K\hat\alpha \\
		\lambda K\hat\alpha &= KY - K^2\hat\alpha \\
		(K + \lambda) K\hat\alpha &= KY \\
		K\hat\alpha &= (K + \lambda)^{-1} KY.
	\end{align*}
Since $K$ and $(K + \lambda)^{-1}$ are simultaneously diagonalisable, they commute, and therefore $K\hat\alpha = K(K + \lambda)^{-1} Y$. 
%\begin{align*}
%(K + \lambda)^{-1} KY &= (U(D + \lambda) U\T)^{-1} UDU\T Y \\
%&= U(D + \lambda)^{-1} D U\T Y \\
%&= UD (D+ \lambda)^{-1} U\T Y\\
%&= UDU\T (U (D+ \lambda) U\T) Y\\
%&= K(K + \lambda)^{-1} Y,
%\end{align*}
%which proves the claim.
\end{solution}

\begin{question}
	...
\end{question}

\begin{proof}
	content...
\end{proof}

\begin{question}
	...
\end{question}

\begin{proof}
	...
\end{proof}

\begin{question}
	Show from first principles that the Sobolev kernel is indeed a (positive definite) kernel.
\end{question}

\begin{solution}
	Let $\XX = [0, 1]$ and $k(x, x') = \min(x, x')$ the Sobolev kernel. We must then show that for any $x_1, \dotsc, x_n \in \XX$, the matrix $K_{ij} = \min(x_i, x_j)$ is positive semi-definite. 
\end{solution}
\end{document}

### 2020-10-20 topics_statistical_theory/C1.tex
updated notes and worked on msm exercises

\begin{proof}
    \begin{enumerate}[(a)]
        \item We first consider $\PP(X \geq x)$. 
        By Markov's inequality, we have for all $t \in \RR$ that
        \[
        \PP(X \geq x) = \PP(e^{tX} \geq e^{tx}) \leq e^{-tX} \EE(e^{tX}) \leq e^{-tx + t^2\sigma^2/2}.  
        \]
        Since the LHS is independent of $t$, we can take the infimum over $t$ on the RHS and obtain
        \[
        \PP(X \geq x) \leq \inf_{t \in \RR} e^{-tx + t^2\sigma^2/2} = e^{-x^2/(2\sigma^2)}, 
        \]
        since the infimum of $t^2\sigma^2/2 - tx$ is attained at $t = x/\sigma^2$.

@@

\begin{theorem}[Hoeffding's inequality]
	Let $X_1, \dotsc, X_n$ be independent sub-Gaussian random variables, with $X_i$ having parameter $\sigma_i^2$. Then $\bar X$ is sub-Gaussian with parameter $\overline{\sigma^2}$. In particular, we have
	\[
	\max\qty{\PP(\bar X \geq x), \PP(\bar X \leq -x)} \leq e^{-nx^2/(2\overline{\sigma^2})}.
	\]
\end{theorem}

@@

>>>>

\begin{proof}
    \begin{enumerate}[(a)]
        \item We first consider $\PP(X \geq x)$. 
        By Markov's inequality, we have for all $t \in \RR$ that
        \[
        \PP(X \geq x) = \PP(e^{tX} \geq e^{tx}) \leq e^{-tX} \EE(e^{tX}) \leq e^{-tx + t^2\sigma^2/2}.  
        \]
        Since the LHS is independent of $t$, we can take the infimum over $t$ on the RHS and obtain
        \[
        \PP(X \geq x) \leq \inf_{t \in \RR} e^{-tx + t^2\sigma^2/2} = e^{-x^2/(2\sigma^2)}, 
        \]
        since the infimum of $t^2\sigma^2/2 - tx$ is attained at $t = x/\sigma^2$ (this method is called \emph{Chernoff bounding}).

@@

\begin{theorem}[Hoeffding's inequality] \label{thm:hoeffding}
	Let $X_1, \dotsc, X_n$ be independent sub-Gaussian random variables, with $X_i$ having parameter $\sigma_i^2$. Then $\bar X$ is sub-Gaussian with parameter $\overline{\sigma^2}$. In particular, we have
	\[
	\max\qty{\PP(\bar X \geq x), \PP(\bar X \leq -x)} \leq e^{-nx^2/(2\overline{\sigma^2})}.
	\]
\end{theorem}

@@

\begin{corollary}
	Let $X_1, \dotsc, X_n$ be independent random variables where $\EE[X_i] = \mu_i$ and $a_i \leq X_i \leq b_i$. Then we have
	\[
	\PP(\bar X - \bar \mu \geq x) \leq \exp\qty(-\frac{2n^2x^2}{\sum_{i=1}^n (b_i - a_i)^2}). 
	\]
\end{corollary}

\begin{proof}
	By Hoeffding's lemma, $X_i - \mu_i$ is sub-Gaussian with parameter $(b_i - a_i)^2/4$ for each $i$. The result now follows from \cref{thm:hoeffding}.
\end{proof}

Note that when $X$ takes values in $[a, b]$, its variance is at most $(b-a)^2$. However, when $\Var(X_i) \ll (b_i - a_i)^2$, Hoeffding's inequality can be loose (for example, when $X_i \sim \Rm{Bern}(p_i)$ with $p_i$ small). In such circumstances, Bennett's or Bernstein's inequality may give better results.

\begin{theorem}[Bennett's inequality]
	Let $X_1, \dotsc, X_n$ be independent random variables with $\EE[X_i] = 0$, $\sigma_i^2 \ceq \Var(X_i) < \infty$, and $X_i \leq b$ for some $b > 0$. Define  $S \ceq \sum_{i=1}^n X_i$, $\nu \ceq \nu$ and $\phi \colon \RR \to \RR$ by $\phi(u) \ceq e^u - 1 - u = \sum_{k=2}^\infty \frac{u^k}{k!}$, then for every $t > 0$ we have
	\[
	\log\EE[e^{tS}] \leq \frac{n \nu }{b^2} \phi(bt). 
	\]

	Defining $h \colon (0, \infty) \to [0, \infty)$ by $h(u) \ceq (1+u)\log(1+u) - u$, we have for every $x > 0$ that
	\[
	\PP(\bar X \geq x) \leq \exp\qty(-\frac{n \nu}{b^2} h\qty(\frac{bx}{\nu})). 
	\]
\end{theorem}

\begin{proof}
	Define $g \colon \RR \to \RR$ by 
	\[
	g(u) \ceq \sum_{k=0}^\infty \frac{u^k}{(k+2)!} = \begin{cases}
		\frac{\phi(u)}{u^2}&\text{if $u \neq 0$}, \\
		\frac12 &\text{if $u = 0$}. 
	\end{cases} 
	\]
	Then one can check that $g$ is increasing on $\RR$, so 
	\[
	e^{tX_i} - 1 - t X_i = t^2 X_i^2 g(tX_i) \leq t^2 X_i^2 g(tb) = X_i^2 \frac{\phi(bt)}{b^2},
	\]
	and therefore
	\[
	e^{tX_i} \leq 1 + tX_i + X_i^2 \frac{\phi(bt)}{b^2} \implies \EE[e^{tX_i}] \leq 1 + \EE[X_i^2] \frac{\phi(bt)}{t^2} = 1 + \Var(X_i) \frac{\phi(bt)}{b^2}. 
	\]

	Hence for $t > 0$ we have
	\begin{align*}
		\log \EE[e^{tS}] &= \sum_{i=1}^n \log\EE[e^{tX_i}] \leq n \cdot \frac1n\sum_{i=1}^n \log\qty(1 + \Var(X_i) \frac{\phi(bt)}{b^2}) \\
		&\overset*\leq n \log(1 + \frac{\nu \phi(bt)}{b^2}) \overset{**}\leq  \frac{n\nu}{b^2}\phi(bt).
	\end{align*}
Here, $(*)$ follows from the fact that $\log$ is a concave function while $(**)$ follows from the fact that $\log(1+u) \leq u$ for all $u \geq 0$. This concludes the proof for the first part of the theorem.

Now, we apply the method of Chernoff bounding and find
\[
\PP(\bar X \geq x) = \PP(S \geq nx) \leq \inf_{t > 0} e^{-ntx} \EE[e^{t S}] \leq \inf_{t > 0} e^{-ntx + n\nu \phi(bt)/b^2} = \exp\qty(-\frac{n\nu}{b^2} h\qty(\frac{bx}{\nu})),
\]
since once can check that the infimum is attained at $t = b^{-1} \log(1 + bx/\nu)$. 
\end{proof}

\begin{definition}
	A random variable $X$ with $\EE X = 0$ is called \emph{sub-Gamma in the right tail} with variance factor $\sigma^2 > 0$ and scale $c > 0$ if
	\[
	\EE[e^{tX}] \leq \exp\qty(\frac{\sigma^2 t^2}{2(1 - ct)})
	\]
	for all $t \in [0, 1/c)$. 
\end{definition}

Note that this definition looks like that of sub-Gaussianity, except that $e^{tX}$ can explode as $t$ approaches $1/c$. We give some characteristics of sub-Gamma distributions:

\begin{definition}
	For any $x \in \RR$ we define $x_+ \ceq \max(x, 0)$. 
\end{definition}

\begin{proposition}
\begin{enumerate}[(a)]
\item Let $X$ be sub-Gamma in the right tail with variance factor $\sigma^2$ and scale $c$. Then 
\[
\PP(X \geq x) \leq \exp\qty(-\frac{x^2}{2(\sigma^2 + cx)})
\]
for all $x \geq 0$.

\item Let $X$ be a random variable with $\EE X = 0$, $\EE[X^2] \leq \sigma^2$ and $\EE[(X_+)^q] \leq q! \sigma^2 c^{q-2}/2$ for all $q \geq 3$. Then $X$ is sub-Gamma in the right tail with variance factor $\sigma^2$ and scale parameter $c$. 
\end{enumerate}
\end{proposition}

\begin{proof}
\begin{enumerate}[(a)]
\item Again, we apply a Chernoff bound: we have
\begin{align*}
\PP(X \geq x) \leq \inf_{t \in [0, 1/c)} e^{-tx} \EE[e^{tX}] &\leq \inf_{t \in [0, 1/c)} \exp(-tx + \frac{\sigma^2t^2}{2(1 - ct)}) \\
&\leq \exp(- \frac{x^2}{2(\sigma^2 + cx)}),
\end{align*}
where we have set $t = x/(\sigma^2 + cx) \in [0, 1/c)$ in the final step.

\item Recall from the proof of Bennett's inequality that $g$ is increasing and therefore for $u \leq 0$ we have $\phi(u) = u^2 g(u) \leq u^2 g(0) = \frac{u^2}{2}$. Therefore, for every $u \in \RR$ we have
\[
\phi(u) \leq \frac{u^2}{2} + \sum_{q=3}^\infty \frac{(u_+)^q}{q!}.
\]
We deduce that for $t \in [0, 1/c)$ we have (note $\log(x) \leq x - 1$ for all $x$):
\[
\log \EE[e^{tX}] \leq \EE(e^{tX}) - 1 = \EE[\phi(tX)] \leq \EE \qty[\frac{t^2X^2}{2} + \sum_{q=3}^\infty \frac{t^q X_+^q}{q!}].
\]
By Fubini's theorem, since the infinite sum has only positive terms we may interchange sum and expectation to obtain
\[
\EE \qty[\frac{t^2X^2}{2} + \sum_{q=3}^\infty \frac{t^q \EE[X_+^q]}{q!}] =  \frac{t^2 \Var[X]}{2} + \sum_{q=3}^\infty \frac{t^q \EE[X_+^q]}{q!} \leq \frac{\sigma^2 t^2}{2} \sum_{q=2}^\infty t^{q-2} c^{q-2} = \frac{\sigma^2t^2}{2(1-ct)}. 
\]
\end{enumerate}
\end{proof}

Following this proposition, we can prove Bernstein's inequality:
\begin{theorem}[Bernstein's inequality]
	Let $X_1, \dotsc, X_n$ be independent random variables with $\EE[X] = 0$, $\frac1n\sum_{i=1}^n \Var(X_i) \leq \sigma^2$ and $\frac1n \sum_{i=1}^n \EE[(X_i)_+^q] \leq q! \sigma^2 c^{q-2}/2$ some $\sigma, c > 0$ and for all $q \geq 3$. Then $S \ceq \sum_{i=1}^n X_i$ is sub-Gamma in the right tail with variance factor $n\sigma^2$ and scale parameter $c$. In particular we have
	\[
	\PP(\bar X \geq x) \leq \exp\qty(- \frac{nx^2}{2(\sigma^2 + cx)}). 
	\]
\end{theorem}

\begin{proof}
	We have by part (b) of the previous proposition
	\[
	\log\EE[e^{tS}] = \sum_{i=1}^n \log \EE[e^{tX_i}] \leq n \frac{\sigma^2 t^2}{2(1 - ct)},
	\]
	and the second claim follows from part (a) of the previous proposition:
	\[
	\PP(\bar X \geq x) = \PP(S \geq nx) \leq \exp\qty(- \frac{nx^2}{2(\sigma^2 + cx)}). 
	\]
\end{proof}
